# Practicum 1.1

## Working directory

```{r}
getwd()
```

## Loading and observing dataset (Task 1, 2 )

```{r}
library(readr)
# data <- read.csv("./players_22.csv")
data <- read_csv("./players_22.csv")
dim(data)
```

#### Column labels (Task 2)

```{r}
colnames(data)
```

#### Heterogoneous or homogeneous data

If a column contains more than one type after the : it is heterogeneous, otherwise its homogeneous.

```{r}
str(data)
```

#### Can a tibble be replaced with a Matrix or Array?

A matrix and array in R require all values to be of the same type. An array is similar but can have multiple dimensions. We can use \*sapply\*

```{r}

is_numeric <- sapply(data, is.numeric)
all(is_numeric)
```

### Occurrences of each column type (task 3)

```{r}

# we get a named character vector of types of each column
column_types <- sapply(data, class) 
# we get the frequency table of each type
type_counts <- table(column_types)    

print(type_counts)
```

### Complete rows (task 4)

```{r}
# complete cases return a logical vector of TRUE/FALSE flags
# that indicate whether row is complete
complete_players <- complete.cases(data)
print(complete_players[0:10])

```

#### Final count

```{r}

# now we get the sum of such rows
print(sum(complete_players))
```

### Duplicates (task 5)

#### Rows

In order to collect duplicate flags we can apply \*duplicated()\* function.

```{r}

# logical vector of TRUE/FALSE to imply the duplicate row
duplicate_rows <- duplicated(data)
print(sum(duplicate_rows))
```

#### Columns

We first transpose the matrix and then use an analogous method to detect duplicate columns.

```{r}

data_transposed <- as.data.frame(t(data))
print(sum(duplicated(data_transposed)))

```

### Value range (task 6)

#### Numeric columns

```{r}
library(dplyr)

numeric_ranges <- data %>%
  select(where(is.numeric)) %>%
  # summarize the rows by applying this list of functions on every column
  summarise(across(everything(), list(min = ~min(.), max = ~max(.))))

# Filter numeric columns and get min and max
# numeric_ranges <- data %>%
#   select(where(is.numeric)) %>%  # Select only numeric columns
#   summarise(across(everything(), list(min = ~min(.), max = ~max(.))))  # Find min and max

# Print the result
print(numeric_ranges)
```

#### Enum

```{r}
library(dplyr)

category_choices <- data %>%
  select(where(is.factor) | where(is.character)) %>%  # Select only factor and character columns
  summarise(across(everything(), ~list(unique(.))))  # Apply unique() and return the result as a list

print(category_choices)

```

### Examining the structure (task 7)

```{r}
View(data)
```

After examining the data, we have found the following information:

-   **Player's personal attributes** (id, name, last_name, age, dob, etc.)

-   **Player's club attributes** (club_name, club_position, etc.)

-   **Player's performance attributes** (pace, shooting, skill curve, mentality_penalties, cm, rw, lw, etc.)

-   **Other attributes** (urls, tags, etc.)

    **Player's performance metrics** are often numeric columns that represent different aspects of the player's ability.

    **Non-informative attributes** include mainly the urls, tags, id, nationality_id, name, long_name, etc.

    #### Reordering the dataset

    The very first columns must be some unique and important identifier, such as short_name. Then some additional personal attributes follow (dob, club_name, club_position, etc.). After these, the performance attributes follow and then the rest. Possible reordering example:

```{r}

library(dplyr)

data_rearranged <- data %>%
  select(short_name, dob, club_name, club_position, overall, shooting, passing, dribbling, everything())

View(data_rearranged)
```

### Player with highest overall (task 8)

```{r}

library(dplyr)

top_player <- data_rearranged %>%
  select(short_name, dob, club_name, club_position, wage_eur, overall) %>%
  filter(overall == max(overall))

print(top_player)
```

### Left-footed players (task 9)

```{r}
library(dplyr)

# Compute total number of players
nrows <- nrow(data_rearranged)

# Compute total left- and right-footed players
left_count <- sum(data_rearranged$preferred_foot == "Left", na.rm = TRUE)
right_count <- sum(data_rearranged$preferred_foot == "Right", na.rm = TRUE)

left_footed <- data_rearranged %>%
  select(short_name, dob, club_name, club_position, overall, preferred_foot) %>%
  filter(preferred_foot == "Left") %>%
  arrange(desc(overall)) %>%
  mutate(
    proportion_left = left_count / nrows,  # Proportion of left-footed players
    proportion_right = right_count / nrows # Proportion of right-footed players
  )

print(left_footed)
```

### Left-footed players overall (task 10)

After comparing the overall of left-foot and right-foot players we can see that left-foot players have slightly greater overall. The counts of each category, however, are not equal.

```{r}
library(dplyr)

avg_overall <- data_rearranged %>%
  filter(preferred_foot %in% c("Left", "Right")) %>%
  group_by(preferred_foot) %>%
  summarise(avg_overall = mean(overall, na.rm = TRUE), count = n())


print(avg_overall)
```

### Youngest left-footed player (task 11)

```{r}
library(dplyr)

youngest_left_foot_pl <- data_rearranged %>%
  select(short_name, dob, club_name, overall, preferred_foot) %>%
  filter(preferred_foot == 'Left' & overall > 80) %>%
  arrange(desc(dob)) %>%
  head(n=1)

print(youngest_left_foot_pl)
```

### Least-valued players with greatest potential

```{r}
library(dplyr)
library(lubridate)  # For handling dates

# Ensure age column exists; if not, compute it from DOB
data_rearranged <- data_rearranged %>%
  mutate(age = ifelse("age" %in% names(.), age, as.numeric(difftime(Sys.Date(), ymd(dob), units = "days") / 365.25)))

# Identify least-valued left-footed players with greatest potential improvement
least_valued_players <- data_rearranged %>%
  filter(preferred_foot == "Left", age < 35) %>%  # Under 35
  arrange(value_eur, desc(potential)) %>%  # Sort by lowest value, highest potential
  select(short_name, age, club_name, overall, potential, value_eur) %>%
  head(n = 5)  # Select top 5

print(least_valued_players)
```

### Other insights (task 13)

#### Highest dribbling skill

```{r}
library(dplyr)

highest_dribbling_skill_pl <- data_rearranged %>%
  select(short_name, age, club_name, preferred_foot, dribbling) %>%
  filter(preferred_foot == 'Left') %>%
  arrange(desc(dribbling)) %>%
  head(n=1)

print(highest_dribbling_skill_pl)
```

#### Amount of left-footed players with shooting attributes over 85

```{r}

library(dplyr)

shooting_attribute_pl <- data_rearranged %>%
  select(short_name, age, club_name, preferred_foot, shooting) %>%
  filter(preferred_foot == 'Left' & shooting > 85) %>%
  arrange(desc(shooting)) %>%
  count()

print(shooting_attribute_pl)
```

#### Best left-player in bending the ball

```{r}

library(dplyr)

best_curving_left_footed <- data_rearranged %>%
  filter(preferred_foot == "Left") %>%
  arrange(desc(skill_curve)) %>%
  select(short_name, age, club_name, overall, skill_curve) %>%
  head(1)  # Get the top player

print(best_curving_left_footed)
```
